---
title: "Danish Kings"
author: "Valdemar Nissen"
date: "2026-03-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The task here is to load your Danish Monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the kings

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `kings.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the separator: Comma  

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}
# FILL IN THE CODE BELOW and review the outputs
library(tidyverse)
kings1 <- read.csv("data/kings.csv")
kings2 <- read_csv("data/kings.csv")
kings3 <- read.csv2("data/kings.csv")
kings4 <- read_csv2( "data/kings.csv")
```
Answer: 
1. Which of these functions is a `tidyverse` function? Read data with it below into a `kings` object
the read_csv and read_csv2 is a tidyverse function

2. What is the result of running `class()` on the `kings` object created with a tidyverse function.
The result of running class(kings) on the kins object created with a tidyverse function shows "spec_tbl_df" "tbl_df"      "tbl" "data.frame". This means that the kings object is a tibble and a dataframe created by the tidyverse

3. How many columns does the object have when created with these different functions?
The object have 11 columns when its read by the command read_csv. Other objects have different answers to how many columns they contain, which can be explained by how the command reads the data and the seperators 

4. Show the dataset so that we can see how R interprets each column
```{r eval = TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

kings <- read_csv("data/kings.csv")

class(kings)

typeof(kings)

ncol(kings)

head(kings)

view(kings)

glimpse(kings)

```
## Calculate the duration of reign for all the kings in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`


Create a new column called  `duration` in the kings dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r}
# YOUR CODE
library(tidyverse)
kings <- kings%>% 
  mutate(
    Start_year=as.numeric(Start_year),
    End_year=as.numeric(End_year),
    duration=End_year-Start_year)
sum(is.na(kings$duration))
# This code is converting Start_year and End_year to numeric values using the Start_year=as.numeric(Start_year) and End_year=as.numeric(End_year). Then the code is creating a new column called duration, where it contains the results of subtracting the End_year with Start_year. The separate code is showing how many missing values (NA) is shown in the new duration column 

```

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `kings` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
My options for choosing the relevant duration column are kings$duration, kings["duration"] or select(kings,duration).

* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.

If the chosen column is selected as kings$duration, the column is a vector.
If the chosen column is seleceted as select(kings,duration) or kings["duration"], the column is a tibble

* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`
```{r}
average_duration_vector <- mean(kings$duration,na.rm=TRUE)
average_duration_tibble <- kings%>% summarize(average_duration=mean(duration,na.rm=TRUE))

  
```

## How many and which kings enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `kings` dataset. Display the result and also  count the resulting rows with `count()`

```{r}
kings_longer_than_average_rule <- kings%>% 
  filter(duration>average_duration_vector)
select(kings_longer_than_average_rule,Name,duration)
count(kings_longer_than_average_rule)
# This code are filters the kings dataset to show what monarch rule duration are longer than the average. The result of the code are now stored in the kings_longer_than_average_rule. The select select(kings_longer_than_average_rule,Name,duration) shows the monarchs and their duration. count(kings_longer_than_average_rule) are counting how many monarchs reign period are longer than the average

```
## How many days did the three longest-ruling monarchs rule? 

* Sort kings by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r}
# YOUR CODE
longest_monarch_rule <- kings%>%
  arrange(desc(duration))%>%
  slice(1:3)%>%
  mutate(Days=duration*365.25)%>%
  select(Name,duration,Days)

# This code takes the "kings" dataset and sorts it by duration in a descending order. The arrange() function sorts the rows and the desc(duration) guarantees that the longest durations appears first. Then it selects the top three results over how long the monarchs ruled with the slice(1:3) function. Further more, the code also creates a new column called Days, where its converting years into days with the mutate function. By typing 365.25 you are also including the transition years. The results are now listed as a object called longest_monarch_rule by the <- command
  
longest_monarch_rule


```

## Challenge: Plot the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r}
# YOUR CODE
library(dplyr)
library(ggplot2)

kings <- kings%>%
  mutate(midyear=Start_year+(End_year-Start_year)/2)

ggplot(
  kings%>%filter(!is.na(duration)&!is.na(midyear)),
  aes(x=midyear,y=duration)
)+
  geom_point(size=3,alpha=0.6,color="steelblue")+
  geom_smooth(method="loess",se=TRUE,color="darkred")+
  labs(
    title="Duration of Reign of Danish Monarchs Over Time",
    x="Midyear of Reign",
    y="Duration of Reign (Years)"
  )+
  theme_bw(base_size = 14)
# The library(dplyr) enables mutate, filter, %>% and (ggplot2) enables ggplot(), geom_point(), geom_smooth()
# First, we create a new column inside the kings data, which is called midyear. midyear are now the middle of the monarchs rule overtime
# With using the ggplot function, we are now creating a graph. By doing the is.na function, we are removing rows where values are missing. The aes(x=midyear,y=duration) function are ensuring the graphs x-axis are midyear and the y-axis are duration
# Then, by doing the geom_point function, we are creating data points on the graph. Size=3 are choosing the size, alpha are choosing the visibility and color choses the color of the data points
# Then, by doing the geom_smooth, we are creating a trend line for the graph. method=loess creates a curve that follows the data. se=TRUE creates a darker area around the trend line, that shows a uncertainty of the the line. Color choses the color of the trend line
# The theme_bw(base_size = 14) gives the graph a theme and a size for the text
```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!




